Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-08-24 paper/ms.tex
Further updates to text

% region preamble
% mnras_template.tex 
%
% LaTeX template for creating an MNRAS paper
%
% v3.0 released 14 May 2015
% (version numbers match those of mnras.cls)
%
% Copyright (C) Royal Astronomical Society 2015
% Authors:
% Keith T. Smith (Royal Astronomical Society)

% Change log
%
% v3.0 May 2015
%    Renamed to match the new package name
%    Version number matches mnras.cls
%    A few minor tweaks to wording
% v1.0 September 2013
%    Beta testing only - never publicly released
%    First version: a simple (ish) template for creating an MNRAS paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
% Depending on your LaTeX fonts installation, you might get better results with one of these:
%\usepackage{mathptmx}
%\usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}

@@

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{newtxtext,newtxmath}

@@

\newcommand{\Nbody}{$N$-body}
\newcommand{\AREPO}{\textsc{AREPO}}
\newcommand{\SMUGGLE}{SMUGGLE}

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared

@@

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
keyword1 -- keyword2 -- keyword3
\end{keywords}

@@

\section{Introduction}
Elongated bar-like features are ubiquitous in galaxies, occurring at the centers
of approximately two-thirds of spiral disks \citep{2000AJ....119..536E,
2007ApJ...657..790M} and in our own Milky Way \citep{1957AJ.....62...19J,
1991ApJ...379..631B}. The ubiquity of bars is not difficult to explain, since
stellar disks simulated in isolation almost always form bar-like structures
\citep{1971ApJ...168..343H}. Several studies have shown that a spherical
potential, such as from a stellar bulge or dark matter halo, acts to stabilize a
stellar disk against the bar instability \citep{1973ApJ...186..467O,
1976AJ.....81...30H}.

@@

While the interaction between the bar and a dark matter halo is well-understood
theoretically, this is not the case for the interaction between a bar and a
gaseous disk. Some argue that the gas disk should slow down the bar more, while
others argue that the tendency of the bar to drive gas inwards means the bar
should speed up due to the effect of the gas disk.

In this work, we present a simulation of an isolated, gas-rich disk galaxy. Our
galaxy exhibits some similarities to the Milky Way, described in more detail in
Section XX. Surprisingly, we find that the pattern speed of the bar in this
galaxy is constant with time. We find that with even modest gas fractions of
about 5\%, our isolated galaxy displays no evolution in pattern speed over about
$8\,\textrm{Gyr}$. This behavior has been observed in a few previous works, but
we are the first to exhibit that the behavior ought to be present in a large
fraction of galaxies. We are also the first to provide a physical explanation of
the behavior with clear observational predictions.

In section 2, \dots

@@

\section{Methods}

@@

@@

\begin{figure}
\centering
\includegraphics[width=9cm]{fig/A2_over_A0.pdf}
\caption{\textbf{Evolution of bar strength.} The bar strength is measured as
the maximum of the second Fourier component divided by the zeroth Fourier
component. We see that in the
\Nbody{} case (blue) the bar strength increases with time, consistent with
previous results showing that the bar strength increases as bars slow down. In
the SMUGGLE case (orange), we see that the bar strength remains roughly
constant, possibly slightly decreasing with time. This is also consistent with
the expected relation between pattern speed and strength since the bar in this
case is not slowing down.}
\label{fig:strength}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=18cm]{fig/bar_decomp.pdf}
\caption{\textbf{Disk decomposition into the barred and unbarred disk.} This
procedure is based on \citet{2016MNRAS.463.1952P} The \textit{left panel} shows
a face-on surface density projection through the stellar component of the
\SMUGGLE{} simulation (disk and bulge) at $t=1\,\textrm{Gyr}$. The
\textit{middle panel} shows the component of the disk identified as being
trapped in the bar while the \textit{right panel} shows the component of the
disk identified as not being trapped in the bar. The fact that the untrapped
stars form a roughly axisymmetric structure indicates our bar decomposition is
sufficiently accurate.}
\label{fig:decomp}
\end{figure*}

The time evolution of the bar strength, defined as the maximum of
$\left|A_2/A_0\right|$ as a function of radius, is shown in Extended Data
Fig.~\ref{fig:strength}. The quantity $\left|A_2/A_0\right|$ varies from $0$
to $1$, with larger values indicating a stronger bar pattern. We see that in
the \Nbody{} case, $\left|A_2/A_0\right|$ increases over time as the bar pattern
slows. This is consistent with previous \Nbody{} simulations which showed a
clear correlation between the bar pattern speed and the bar
strength\cite{2003MNRAS.341.1179A}. In the SMUGGLE case, we see that the bar
strength has an initial drop but then remains at a roughly constant, but
slightly decreasing, strength. This is consistent with the pattern speed in the
SMUGGLE case being roughly constant or slightly increasing.

Computing the length of the bar and the torque on the bar by different
components requires us to decompose the disk into a component which is trapped
by the bar and a component which is untrapped. In order to do this, we follow
closely the technique developed in ref.\cite{2016MNRAS.463.1952P} We analyzed
the orbit of each star particle (meaning initial disk, bulge, and newly formed
stars) by extracting the $x$-$y$ positions of the apoapse of each in a frame
corotating with the bar (apoapses are defined as local maxima in $r$). For
each apoapse, we searched for the $19$ closest apoapses in time and applied a
$k$-means clustering algorithm on this set of $20$ points with $k=2$. We then
computed for each of the two clusters the average angle from the bar
$\left<\Delta \phi\right>_{0,1}$, the standard deviation in $R$ of the points
${\sigma_R}_{0,1}$, and the average radius of the cluster
$\left<R\right>_{0,1}$. At each apoapse, a particle was considered to be in
the bar if it met the following criteria:
\begin{equation}
\textrm{max}\left(\left<\Delta \phi\right>_{0,1}\right) < \pi / 8
\end{equation}
\begin{equation}
\frac{{\sigma_R}_0 + {\sigma_R}_1}{\left<R\right>_0 + \left<R\right>_1} < 0.22
\end{equation}
These criterion are slightly different and simplified from the ones used in
ref.\cite{2016MNRAS.463.1952P}, but we found them to empirically work well at
decomposing the disk into a bar and disk component. In Extended Data
Fig.~\ref{fig:decomp}, we show an example of this decomposition. The
\textit{left} panel shows a surface density projection of the stellar disk and
bulge (including newly formed stars) from the SMUGGLE model after
$1\,\text{Gyr}$ of evolution in a frame such that the bar is aligned with the
$x$-axis. The \textit{middle} panel shows a projection of the subset of stars
that are identified as being trapped in the bar and the \textit{right} panel
shows a projection of the stars that are not identified as being trapped. The
fact that the \textit{right} panel is roughly axisymmetric indicates the bar
decomposition is performing adequately.

After the disk has been decomposed into a trapped and untrapped component, we
measured the bar length as being the radius $R_b$ which encapsulates $99\%$ of
the stars identified as being trapped in the bar, allowing for some outliers.
For the computation of the torque, we used the tree algorithm in
\texttt{MakeNewDisk}\cite{2005MNRAS.361..776S} customized to be accessible
from \texttt{Python} using \texttt{Cython}. This algorithm is based on the
\texttt{TREESPH} code\cite{1989ApJS...70..419H}. We constructed a tree with an opening angle of $0.35$ using
only the star particles identified as being trapped in the bar. We then queried the tree at the locations of all
resolution elements in the other components and computed the torque of the bar
on such components. The torque on the bar by the other components is simply
the negative of the torque on the other components by the bar. A similar
method was applied in measuring the torque for the disk when the pattern speed
is kept constant, as is done in Fig.~\ref{fig:equil}.

% \subsection{Idealized Mock Light Images}
% We created synthetic mock light images at $1024^2$ resolution in
% Fig.~\ref{fig:overview} for the \textit{Hubble Space Telescope} using a
% post-processed Monte Carlo radiation transfer simulation with the SKIRT9 code
% including secondary emission from dust\cite{2020AC....3100381C}. Emission from
% star particles was calculated assuming a Bruzual-Charlot spectral energy
% distribution\cite{2003MNRAS.344.1000B} and a Chabrier initial stellar mass
% function\cite{2003PASP..115..763C}. The luminosity from each star particle was
% smoothed using a cubic spline kernel with each star particle being assigned a
% smoothing length equal to its distance to its $32$nd closest neighbor. We
% launched $10^9$ photon packets per simulation segment. For star particles
% which were present in the initial simulation (i.e., initial disk and bulge
% particles), we assumed that they have an age of $5\,\textrm{Gyr}$ and solar
% metallicity. For star particles in the SMUGGLE simulation which are newly
% formed, we used their recorded ages and metallicities.

% For the SMUGGLE simulations, we included absorption and emission from dust. We
% assumed the THEMIS (The Heterogeneous dust Evolution Model for Interstellar
% Solids) dust model\cite{2017AA...602A..46J} to convert gas metallicity and
% mass density to dust mass density. For the \Nbody{} simulations we ran the
% SKIRT9 code in a mode without an obscuring medium, but otherwise applied the
% same procedure.

% We used the profiles for filters F814W, F606W, and F475W on the Advanced
% Camera for Surveys (ACS) instrument for red, green, and blue pixel values,
% respectively. Filtered fluxes were converted to RGB values using an $\arcsinh$
% scaling following ref.\cite{2004PASP..116..133L} Scalings and cutoff values
% were adjusted manually to create visually appealing images. Filter profiles
% were provided by the Spanish Virtual Observatory Filter
% Service\cite{2012ivoa.rept.1015R, 2020sea..confE.182R}.

@@

\section{Discussion}
Our setup is initially out of equilibrium, but we found that after
$\sim500\,\textrm{Myr}$, the entire system has settled into a steady-state
configuration and initial transients appear not to affect the results after
this point. The constant surface density of the initial gas disk is important
for ensuring the gas disk is dense enough in order for comparisons to real
galaxies to be appropriate.

We also show the evolution of the surface density profile in Extended Data
Fig.~\ref{fig:surf} We find that in our simulation the atomic and molecular
gas surface density and the SFR surface density is broadly consistent with the
expected values for the Milky
Way\cite{2008AA...487..951K,2022ApJ...929L..18E}. The discrepancy between $1$
and $4\,\textrm{kpc}$ in the molecular and SFR surface density is likely due
to the fact that the distances to molecular clouds which underlines this work
used a simple kinematic distance based on an axisymmetric model of the Milky
Way\cite{2017ApJ...834...57M}, which is not accurate in the bar region where
gas has large non-circular velocities.

The pattern speed evolution of this setup is given in Extended Data
Fig.~\ref{fig:GFM}, which shows that the pattern speed evolves in a similar
manner to the SMUGGLE case shown in Fig.~\ref{fig:prop}.

@@

\section{Conclusions}

The last numbered section should briefly summarise what has been done, and describe
the final conclusions which the authors draw from their work.

\section*{Acknowledgements}

The Acknowledgements section is not numbered. Here you can thank helpful
colleagues, acknowledge funding agencies, telescopes and facilities used etc.
Try to keep it short.

>>>>

% region preamble

\documentclass[fleqn,usenatbib]{mnras}

\usepackage[T1]{fontenc}

@@

\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{newtxtext,newtxmath}

@@

\newcommand{\Nbody}{$N$-body}
\newcommand{\AREPO}{\textsc{AREPO}}
\newcommand{\SMUGGLE}{SMUGGLE}
\newcommand{\Msun}{\ensuremath{M_{\odot}}}

\DeclareMathOperator\arcsinh{arcsinh}

@@

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
galaxies: bar - galaxies: disc - galaxies: evolution - galaxies: kinematics and
dynamics
\end{keywords}

@@

\section{Introduction}
Approximately two-thirds of spiral disks host an elongated bar-like feature at
their center \citep{2000AJ....119..536E, 2007ApJ...657..790M}, including our own
Milky Way \citep{1957AJ.....62...19J, 1991ApJ...379..631B}. The ubiquity of bars
is not difficult to explain, since stellar disks simulated in isolation almost
always form bar-like structures \citep{1971ApJ...168..343H}. Several studies
have shown that a spherical potential, such as a stellar bulge or dark matter
halo, acts to stabilize a stellar disk against the bar instability
\citep{1973ApJ...186..467O, 1976AJ.....81...30H}.

@@

While the interaction between the bar and a dark matter halo is well-understood
theoretically, this is not the case for the interaction between a bar and a
gaseous disk. Some argue that the gas disk should slow down the bar more
\citep{2003MNRAS.341.1179A}, while others argue that the tendency of the bar to
drive gas inwards means the bar should speed up due to the effect of the gas
disk \citep{2013MNRAS.429.1949A, 2014MNRAS.438L..81A}. Since the gas phase
typically only contributes about $10-20\%$ of the mass of a galaxy at the
present day, one might naively expect it to have a subdominant effect on the
bar. However, because gas is collisional, it can participate in non-resonant
angular momentum exchange with the bar \citep{2011MNRAS.415.1027H}. Thus,
numerical work has shown that the gas phase can have a stronger influence on a
bar than its contribution to the mass of a galaxy would suggest
\citep{2010ApJ...719.1470V, 2013MNRAS.429.1949A}.

We have performed a simulation of a disk galaxy using the finite-volume,
gravito-hydrodynamics code AREPO \citep{2010MNRAS.401..791S}. We use the galaxy
formation code Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE). This disk galaxy
exhibits almost no evolution in the bar pattern speed over several Gyr when the
gas phase is accounted for and properly modeled. This behavior has been observed
in a few previous works \citep{1993AA...268...65F, 2010ApJ...719.1470V,
2014MNRAS.438L..81A}, but we are the first to exhibit that the behavior ought to
be present in a large fraction of galaxies. We are also the first to provide a
physical explanation of the behavior with clear observational predictions. We
show synthetic \textit{Hubble Space Telescope} images of our barred galaxy in a
case with and without gas in Fig.~\ref{fig:overview}. We see that the bar grows
longer and stronger without gas, while it remains at approximately the same
length and strength when gas is included.

% In this work, we present a simulation of an isolated, gas-rich disk galaxy. Our
% galaxy exhibits some similarities to the Milky Way, described in more detail in
% Section XX. Surprisingly, we find that the pattern speed of the bar in this
% galaxy is constant with time. We find that with even modest gas fractions of
% about 5\%, our isolated galaxy displays no evolution in pattern speed over many
% $\textrm{Gyr}$.

In Section~\ref{sec:methods}, we describe our initial setup, numerical model,
and details on our bar analysis procedures. In Section~\ref{sec:results}, we
summarize the main results from our findings. We discuss these findings at more
length and in the context of previous research in Section~\ref{sec:discussion}
before concluding in Section~\ref{sec:conclusions}.

\begin{figure*}
    \centering
    \includegraphics[width=18cm]{fig/mockHST.pdf}
    \caption{Synthetic Hubble Space Telescope images of our simulations with and
    without the interstellar medium. The upper panels show an \Nbody{} only
    simulation while the lower panels show a simulation which includes the
    SMUGGLE model for the interstellar medium. Each panel is $20\,\textrm{kpc}$
    to a side. Columns show different points in time, separated by
    $1\,\textrm{Gyr}$. We can see that in the \Nbody{} run, the bar grows in
    length and strength. In the SMUGGLE run, the bar remains at approximately
    the same length and strength over the course of the simulation. The \Nbody{}
    model is identical to the GALAKOS model, discussed in the text. Details on
    the production of these synthetic images is given in Appendix~\ref{app:hst}.}\label{fig:overview}
\end{figure*}

@@

\section{Methods}
\label{sec:methods}

@@

Our setup is initially out of equilibrium, but we found that after about
$500\,\textrm{Myr}$, the system has settled into a steady-state configuration
and initial transients appear not to affect the results after this point. The
constant surface density of the initial gas disk is important for ensuring the
gas disk is dense enough in order for comparisons to real galaxies to be
appropriate.

@@

In order to compute other properties of the bar, it is necessary to decompose
the disk into a barred and unbarred component. We achieved this by following
closely the methods described in \citet{2016MNRAS.463.1952P}. Our implementation
is described in more detail in Appendix~\ref{app:bardecomp}. After the disk has
been decomposed into a trapped and untrapped component, we measured the bar
length as being the radius $R_b$ which encapsulates $99\%$ of the stars
identified as being trapped in the bar, allowing for some outliers.

To compute torques we used the tree algorithm in \texttt{MakeNewDisk}
\citep{2005MNRAS.361..776S} customized to be accessible from \texttt{Python}
using \texttt{Cython}. This algorithm is based on the \texttt{TREESPH} code
\citep{1989ApJS...70..419H}. We constructed a tree with an opening angle of
$0.35$ using only the star particles identified as being trapped in the bar. We
then queried the tree at the locations of all resolution elements in the other
components and computed the torque of the bar on such components. The torque on
the bar by the other components is simply the negative of the torque on the
other components by the bar. A similar method was applied in measuring the
torque for the disk when the pattern speed is kept constant, as is done in
Fig.~\ref{fig:equil}.

@@

\section{Results}
\label{sec:results}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/ps_Rbar_torque.pdf}
    \caption{The \textit{upper panel} shows the evolution
    of the pattern speed. As expected, the bar in the \Nbody{} run slows down due
    to interactions between the bar and the dark matter halo. However, the bar in
    the SMUGGLE run does not slow down and instead remains at a constant pattern
    speed. The \textit{middle panel} shows the evolution of the bar length. In the
    \Nbody{} case, the bar lengthens. This occurs because as the pattern speed
    drops, bar-like orbits at larger radii are possible. Stars are captured on
    these orbits, lengthening the bar. This process does not occur in the SMUGGLE
    cases since the bar pattern speed is not decreasing, and therefore the bar
    length remains constant. The \textit{lower panel} shows the torque on the bar
    by different components. The solid lines show the torque exerted by the halo
    in both the \Nbody{} and SMUGGLE cases. The dashed line shows the torque
    exerted by the gas phase in the SMUGGLE run (there is no gas in the \Nbody{}
    run). Details on how these properties are calculated is given in the Methods
    section.}\label{fig:prop}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=9cm]{fig/A2_over_A0.pdf}
    \caption{The evolution in bar strength. The bar strength is measured as the
    maximum of the second Fourier component divided by the zeroth Fourier
    component. We see that in the \Nbody{} case (blue) the bar strength
    increases with time, consistent with previous results showing that the bar
    strength increases as bars slow down. In the SMUGGLE case (orange), we see
    that the bar strength remains roughly constant, possibly slightly decreasing
    with time. This is also consistent with the expected relation between
    pattern speed and strength since the bar in this case is not slowing down.}
    \label{fig:strength}
\end{figure}

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the \Nbody{} (blue)
and SMUGGLE (orange) runs. The pattern speed in the \Nbody{} case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the \Nbody{} case is consistent with a long
line of numerical research on bars in \Nbody{} simulations
\citep{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
2009ApJ...697..293D}.

However, in the SMUGGLE case the pattern speed remains constant. After the first
Gyr of evolution, we find that the pattern speed increases by only $\sim10\%$
over the next $4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the
pattern speed for the \Nbody{} run over the same interval. As we saw
qualitatively in Fig.~\ref{fig:overview}, the middle panel of
Fig.~\ref{fig:prop} shows that the length of the bar in the \Nbody{} case grows
over time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation increases
\citep{2000ApJ...543..704D, 2003MNRAS.341.1179A}.

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the dark
matter halo, whereas the dashed line indicates the torque on the bar by the gas
phase. In the \Nbody{} case, the halo exerts a steady negative torque on the
bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in units of
$10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the SMUGGLE case
exerts a similar negative torque on the bar in the first Gyr of evolution, but
after that the halo exerts a much smaller torque on the bar, averaging only
$-7.8$ in the same units and over the same time interval. The gas in the SMUGGLE
case exerts a steady positive torque averaging $11.7$ over $1\,\textrm{Gyr}$ in
the same units.

The time evolution of the bar strength, defined as the maximum of
$\left|A_2/A_0\right|$ as a function of radius, is shown in
Fig.~\ref{fig:strength}. The quantity $\left|A_2/A_0\right|$ varies from $0$ to
$1$, with larger values indicating a stronger bar pattern. We see that in the
\Nbody{} case, $\left|A_2/A_0\right|$ increases over time as the bar pattern
slows. This is consistent with previous \Nbody{} simulations which showed a
clear correlation between the bar pattern speed and the bar strength
\citep[e.g.,][]{2003MNRAS.341.1179A}. In the SMUGGLE case, we see that the bar
strength has an initial drop but then remains at a roughly constant, but
slightly decreasing, strength. This is consistent with the pattern speed in the
SMUGGLE case being roughly constant or slightly increasing.

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Pattern Speed Evolution}
The lack of evolution in the pattern speed of the SMUGGLE case is intimately
tied to the sudden decrease in torque exerted on the bar by the dark matter
halo. We argue that this can be understood in terms of the halo wake mechanism.
In the \Nbody{} case, halo material which is resonant with the bar will form a
wake which lags behind and exerts a negative torque on the bar, slowing it down
\citep{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}.\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its pattern
speed. However, the bar seems to empirically respond to a negative torque
induced by a halo wake by slowing down.} As the bar slows down, the location of
the resonances in phase space changes, allowing halo material newly resonant
with the bar to participate in forming the wake. However, the gas is a reliable
source of positive torque on the bar, speeding the bar up. This stops the
resonance location from changing such that the halo cannot reinforce the wake,
arresting the process by which the halo can slow the bar down. We term this
process ``exhaustion.''

\begin{figure*}
    \centering
    \includegraphics[width=18cm]{fig/halo_wake.pdf}
    \caption{The wake excited in the dark matter halo. The dark matter halo wake
    is shown in the \Nbody{} case (\textit{middle panel}) and SMUGGLE case
    (\textit{right panel}) after $2.6\,\textrm{Gyr}$ of evolution. The
    \textit{middle} and \textit{right panels} show a surface density projection
    in the $x$-$y$ plane of the dark matter halo after an axisymmetric average
    has been subtracted. The solid line indicates the direction of the bar while
    the dashed line indicates the direction of the halo wake (both measured by
    taking the second Fourier component within a sphere of all material within a
    radius of $4\,\textrm{kpc}$). The \textit{left panel} shows the time
    evolution of the angle difference between the bar and the halo wake, as
    measured from the second Fourier component. After the first Gyr, the angle
    difference in the SMUGGLE case is smaller than in the \Nbody{} case by about
    a factor of two, reflecting how the dark matter halo in the SMUGGLE case is
    unable to exert as negative a torque on the bar as in the \Nbody{} case.}
    \label{fig:wake}
\end{figure*}

We can test this interpretation by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the left panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the \Nbody{} case than in the SMUGGLE case by about a
factor of two. The center and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the \Nbody{} (center) and SMUGGLE
(right) cases at one point in time.

The presence of the gas can arrest the process by which additional material in
the dark matter halo can contribute to a wake. However, this does not explain
why the pattern speed in the SMUGGLE case is nearly constant over several Gyr.
Naively, it would be a coincidence that the bar pattern speed remains constant
in the SMUGGLE case, resulting from a chance cancellation of the halo and gas
torques. However, a constant pattern speed in the presence of gas has been
observed in a few simulations of barred galaxies with
gas \citep{1993AA...268...65F, 2010ApJ...719.1470V, 2014MNRAS.438L..81A}. Previous
work has argued this is due to the bar torquing gas inwards, but no explanation
has been given for why it might remain constant.

We propose that an equilibrium mechanism is responsible for the pattern speed
remaining approximately constant. In this scenario, residual negative torque
from the dark matter halo balances out the positive torque from the gas phase.
It has been shown when an analytic bar is forced to rotate at a constant pattern
speed for a few Gyr, the halo exerts almost no torque on the bar
\citep{2022MNRAS.513..768C}. We saw in Fig.~\ref{fig:prop} that the dark matter
halo in our simulation is still able to support some negative torque over a
several Gyr time span.

We argue that the following occurs. First, the bar is not able to slow down
quickly enough due to the positive torque of the infalling gas. This causes
the resonant halo phase space at a particular pattern speed, $\Omega_{p,0}$,
to become exhausted and no longer able to support a negative torque. Second,
the gas is still exerting a positive torque on the bar, and therefore the
pattern speed will again increase. Since at higher pattern speeds the halo has
not yet been totally exhausted, the halo will once again be able to exert a
negative torque on the bar. The pattern speed will then settle at a new value
slightly higher than $\Omega_{p,0}$ where the gas and halo torques cancel.
Over time, the pattern speed should slowly increase as the halo becomes
progressively exhausted.

\subsection{Delayed Gas Injection}
A clear prediction of our proposed mechanism is that the constant pattern
speed a particular galaxy will end up is somewhat arbitrary. In the real
universe for an isolated galaxy it would be the formation pattern speed of the
bar while in our simulation it is the pattern speed of the bar when gas is
added to the system. We tested this by adding gas to the system at a later
time when the bar has further grown and slowed down with time. In our
particular test, we added the gas at a time when the pattern speed is
$\sim30\,\textrm{km}/\textrm{s}/\textrm{kpc}$. As shown in
Fig.~\ref{fig:snap700}, we find that the pattern speed evolution is very
similar between the two cases (orange and red lines). If anything, the system
with a lower pattern speed seems to speed up more, which is consistent with
our picture since the stronger bar should experience a larger torque from the
gas as it is more efficient at driving gas inflows. We also show in the
Methods that more slowly rotating bars at fixed bar strength are more
efficient at driving gas inflows as well. Nonetheless, when the initial pattern
speed is lower (red line), the addition of gas does not cause the pattern
speed to quickly return to the higher value of our fiducial simulation (orange
line).

\begin{figure}[t!]%
    \centering
    \includegraphics[width=9cm]{fig/ps_late_start.pdf}
    \caption{Pattern speed evolution with a lower initial pattern speed. We
    tested the evolution of our system when gas is added to the \Nbody{} run at
    a later time, but with all other simulation parameters kept the same. The
    setup is therefore identical to our previous runs just with a bar that is
    larger, stronger, and with a lower pattern speed. We find that the pattern
    speed evolution is very similar to our fiducial case, except that the bar
    retains its original pattern speed. This indicates a mechanism which keeps
    the bar at its formation pattern speed, and that there is not a particular
    pattern speed which the system tends to.}
    \label{fig:snap700}
\end{figure}

\subsection{Varying Initial Gas Fractions}
We also performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Extended Data Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{kpc}^2$ slows down at a similar rate to the \Nbody{} case.

As a result, we conclude that for the disk, bar, and halo properties
considered in this work, a gas fraction of only approximately $5\%$ is
necessary in order for the proposed stabilizing mechanism to operate.

\begin{figure}[t!]
    \centering
    \includegraphics[width=9cm]{fig/ps_fgas.pdf}
    \caption{Pattern speed evolution with varying gas fractions. We explored the
    impact of lowering the initial gas surface density of our fiducial disk on
    the evolution of the pattern speed. The surface densities we tested of $20$,
    $15$, $10$, and $5\,\Msun/\textrm{kpc}^2$ correspond to initial gas
    fractions of $16\%$, $10\%$, $7\%$, and $4\%$. We find that initial surface
    densities $20$, $15$, and $10\,\Msun/\textrm{kpc}^2$ result in bars which
    remain at a constant pattern speed, while an initial surface density of
    $5\,\Msun/\textrm{kpc}^2$ results in a bar which slows down in roughly the
    same manner as the \Nbody{} case.}
    \label{fig:fgas}
\end{figure}

\subsection{Smooth Interstellar Medium}
We also performed a simulation of the same disk but with a simpler model of the
interstellar medium \citep{2003MNRAS.339..289S}, closer to standard methods used
in cosmological simulations of galaxy formation. We find that the pattern speed
evolution is nearly the same in this case, and so conclude that our result is
not sensitive to the details of the model for the interstellar medium. Details
on the physics of this test is given in Section.~X.

\begin{figure}[h!]%
    \centering
    \includegraphics[width=9cm]{fig/ps_GFM.pdf}
    \caption{Pattern speed evolution of a smooth ISM model. This evolution is
    shown for the fiducial disk in the \Nbody{} (blue), SMUGGLE (orange), and
    smooth ISM (red) cases. The smooth ISM model is an older model for the ISM
    which treats its multiphase nature in a subgrid fashion
    \citep{2003MNRAS.339..289S}. This fundamentally differs from the SMUGGLE
    model, which explicitly resolves the hot and cold phases of the ISM
    \citep{2019MNRAS.489.4233M}. The pattern speed in the smooth ISM case is
    broadly similar to the evolution in the SMUGGLE case. This shows that the
    stability of the pattern speed is not simply a result of our assumed model
    for the ISM.}
\label{fig:GFM}
\end{figure}

\subsection{Observations}
We note that ref.\cite{2020MNRAS.491.3655G} found that the rotation parameter
\Rot{} positively correlates with gas fraction, such that galaxies with higher
gas fractions are rotating more slowly. However, it is not obvious this is in
tension with our result since the gas fraction of galaxies correlates with
other galactic properties\cite{2009ARAA..47..159B}. Furthermore, the
measurement of pattern speeds is a delicate process still prone to large error
bars.

\subsection{Cosmological Simulations}
Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations by producing bars which rotate too
slowly\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P, 2021AA...650L..16F}.
However, the pattern speeds of bars in both cosmological simulations and the
real universe can be affected by environmental processes not included in our
simulation -- e.g., satellite infall\cite{2011Natur.477..301P},
non-sphericity\cite{2013MNRAS.429.1949A} or rotation\cite{2013MNRAS.434.1287S,
2014ApJ...783L..18L, 2018MNRAS.476.1331C, 2019MNRAS.488.5788C} in the dark
matter halo, or perhaps even the gaseous circumgalactic medium. Naturally,
extending our present work to account for such effects is a crucial next step
in understanding the formation and evolution of galactic bars.

\section{Conclusions}
\label{sec:conclusions}
The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R}. Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C}. Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence\cite{2019MNRAS.490.4740B}. And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628\cite{2009AA...499L..25C}. This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.
